### 2022-08-02 chapters/optimization/java-dir/java.tex
add : extension to RPC
fix: Proofreading Java chapter

\section*{JAVA}
\section{Introduction}
As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.%ROMAIN: Add source
Furthermore, if we take into consideration legacy applications, Java becomes the most used programming langues.%ROMAIN: Add source
In addition to its popularity, Java exhibits an interesting behaviour when it comes to energy consumption and performances, basically Java applications can be at the same time one of the most energy efficient or hungry solution.
As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
Therefore, we wanted to dig deeper into this aspect of Java and study its runtime.
This chapter thus focuses on the impact of the runtime of Java applications on the energy consumption.

@@

\noindent\textbf{Software Settings.}
For the sake of reproducibility, each experiment runs within a Docker container based on SDKMAN image and Alpline docker.

@@

\vspace{6pt}
\noindent\textbf{Java Virtual Machines Candidates.}
We considered a set of $52$ JVM distributions taken from $8$ different providers/packagers mostly obtained from \textsf{SDKMAN!},\footnote{\url{https://sdkman.io/}} as listed in \Cref{tab:JVM}.
Depending on providers, either all the versions, majors, or LTS are made available by \textsf{SDKMAN!}.

@@

\subsection{Metrics and measurement}
Since the goal of this study is the green aspect Java virtual machines, our key metric will be the energy consumption of the job completed by each configuration.
In addition to the energy consumption, we collected additional metrics to explain the reasons behind the behaviour of each experiment.
Those extra metrics are:
\begin{itemize}
    \item execution time,
    \item number of threads.
\end{itemize}

@@

@@

Then, Figure~\ref{hotspot8} depicts the evolution of the energy consumption of the $12$ benchmarks, when executed on the \textsc{HotSpot} JVM.
\Cref{fig:hotspot8} reports on the energy consumption variation of individual benchmarks, using to \textsc{HotSpot-8}\ as the baseline.
Our results show that the JVM version can severely impact the energy consumption of the application.
However, unlike \Cref{fig:JVMs}, one can observe that, depending on applications, latest JVM versions can consume less energy (60\% less energy for \textsf{Scrabble}) or more energy (25\% more energy for the \textsf{Neo4J}).
It is worth noticing that the energy consumption of some benchmarks, such as \textsf{Reactors}, exhibit large variations across JVM versions due to experimental features and changes that are not always kept when releasing LTS versions (version 11 here).
For example, the introduction of \texttt{VarHandle} to allow low-level access to the memory order modes available in JDK 9 and work along \texttt{Unsafe Classe} that was removed from from JVM 11.\footnote{\url{https://blogs.oracle.com/javamagazine/the-unsafe-class-unsafe-at-any-speed}}

@@

The second interesting situation was observed on the \textsf{Dotty} benchmark.
More specifically, during the first 100 seconds of the execution of the \textsf{Dotty} benchmark on all evaluated JVMs.
At the beginning of the execution, \textsc{GraalVM} has a slightly lower power consumption, is faster, and consumes 10\% less energy.
After about 150 seconds, the power differences between the 3 JVMs is barely noticeable.
One can, however, notice the effect of the JIT, as \textsc{HotSpot} takes the advantage over \textsc{GraalVM} and becomes more energy efficient.
In total, \textsc{HotSpot} completes $597$ requests against $510$ for \textsc{GraalVM} and $381$ for \textsc{J9}, as shown in \Cref{table:power/request}.
\textsc{HotSpot} was thus the best choice on the long term, which explains why it is always necessary to consider a warm-up phase and wait for the JIT to be triggered before evaluating the effect of the JVM or the performance of an application.
This is exactly what we did in our experiments, and why \textsc{HotSpot} was more energy efficient than \textsc{GraalVM} in \Cref{fig:all_benchs}, thus ignoring the warm-up phase would have misleading.

@@

Another measurement issue is the CPU energy variation within machines~\cite{opaper}, thus we executed all the comparable tests on the same node and with the recommended settings to mitigate this threat.

@@

\section{Tools and contributions}
\Cref{table:j-referral} illustrates an example of the final report of \textsf{J-Referral}.
The tool was tested for 2 real Java projects: \textsf{Zip4J} and\footnote{\url{https://github.com/srikanth-lingala/zip4j}} and \textsf{K-nucleotide}.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/performance/knucleotide.html}}.
\textsf{Zip4J} runs a large file compression, while \textsf{K-nucleotide}  extracts a DNA sequence, and updates a hashtable of k-nucleotide keys to count specific values.
The short report presented in \Cref{table:j-referral} shows the ratio of potential energy saving between the most and least energy consuming tested JVM (40\% and 70\% energy savings for \textsf{Zip4J} and \textsf{K-nucleotide} respectively).
Options are available for \textsf{J-Referral} to obtain much more detailed reports including execution time, DRAM usage, split DRAM vs. CPU consumption, etc.
The tool is available as \emph{open-source software} (OSS) from our anonymous repository.\footnote{\url{https://anonymous.4open.science/r/jreferral/Readme.md}}

>>>>

% \section*{JAVA}
\section{Introduction}
As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.%ROMAIN: Add source
Furthermore, if we take into consideration legacy applications, Java becomes the most used programming langues.%ROMAIN: Add source
In addition to its popularity, Java exhibits an interesting behavior when it comes to energy consumption and performances, basically Java applications can be at the same time one of the most energy efficient or hungry solution.
As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
Therefore, we wanted to dig deeper into this aspect of Java and study its runtime.
This chapter thus focuses on the impact of the runtime of Java applications on the energy consumption.

@@

\noindent\textbf{Software Settings.}
For the sake of reproducibility, each experiment runs within a Docker container based on \textsf{SDKMAN!}\footnote{\url{https://sdkman.io}} image and Alpine docker.\footnote{\url{https://github.com/alpinelinux/docker-alpine}}

@@

\vspace{6pt}
\noindent\textbf{Java Virtual Machines Candidates.}
We considered a set of $52$ JVM distributions taken from $8$ different providers/packagers mostly obtained from \textsf{SDKMAN!}, as listed in \Cref{tab:JVM}.
Depending on providers, either all the versions, majors, or LTS are made available by \textsf{SDKMAN!}.

@@

\Cref{fig:jvm-benchmarks-classification} highlights the scope of each benchmark from the test suite.

\begin{figure*}%[ht]
    \centering
    \includegraphics[width=\linewidth]{imgs/jvm_benchmarks}
    \caption{Target scope of \textsc{Dacapo} and \textsc{Renaissance} benchmarks.}
    \label{fig:jvm-benchmarks-classification}
\end{figure*}

\subsection{Metrics and Measurement}
Since the goal of this study is the green aspect of JVM, our key metric will be the energy consumed by job completed for each JVM configuration.
In addition to the energy consumption, we collected additional metrics to explain the reasons behind the behavior of each experiment.
Those additional metrics are:
\begin{itemize}
    \item execution time,
    \item number of threads.
\end{itemize}

@@

\subsection{Extension}
We also added an extension to the protocol to allow the user to run the same experiment with different configurations.
The package is available in the GitHub repository.\footnote{\url{https://github.com/chakib-belgaid/jvm-comparaison}}
To add extra \textbf{jvm candidates} for the benchmark applications, we added a new configuration file \texttt{jvm.sh} in the root directory of the repository, where we put the name and the version of the jvm to be used.\note{The jvm must be provided by \textsf{SDKMAN!}}
For the \textbf{input workload},the benchmarks should be provided in the \texttt{benchmarks} directory.
As for extra \textbf{metrics}, one can create a new script file that monitor the experiment and record the metrics.
We provide some examples, such as \texttt{recordpower.sh} to measure the instantaneous power and \texttt{recordthreads.sh} to measure the number of active threads during the experiment.

For faster experiments, we propose \textsc{Jreferral}\footnote{\url{https://github.com/chakib-belgaid/jreferral}}, an open source tool that automatically compares the JVM energy consumption and recommends the most energy-efficient configuration for a given Java application.
We further discuss this tool in Section~\ref{section:SummaryofContributions}.

@@

Then, \Cref{fig:hotspot8} depicts the evolution of the energy consumption of the $12$ benchmarks, when executed on the \textsc{HotSpot} JVM.
\Cref{fig:hotspot8} reports on the energy consumption variation of individual benchmarks, using to \textsc{HotSpot-8}\ as the baseline.
Our results show that the JVM version can severely impact the energy consumption of the application.
However, unlike \Cref{fig:JVMs}, one can observe that, depending on applications, latest JVM versions can consume less energy (60\% less energy for \textsf{Scrabble}) or more energy (25\% more energy for the \textsf{Neo4J}).
It is worth noticing that the energy consumption of some benchmarks, such as \textsf{Reactors}, exhibit large variations across JVM versions due to experimental features and changes that are not always kept when releasing LTS versions (version 11 here).
For example, the introduction of \texttt{VarHandle} to allow low-level access to the memory order modes available in JDK 9 and work along \texttt{Unsafe Class} that was removed from from JVM 11.\footnote{\url{https://blogs.oracle.com/javamagazine/the-unsafe-class-unsafe-at-any-speed}}

@@

The second interesting situation was observed on the \textsf{Dotty} benchmark.
More specifically, during the first $100$ seconds of the execution of the \textsf{Dotty} benchmark on all evaluated JVMs.
At the beginning of the execution, \textsc{GraalVM} has a slightly lower power consumption, is faster, and consumes 10\% less energy.
After about $150$ seconds, the power differences between the 3 JVMs is barely noticeable.
One can, however, notice the effect of the JIT, as \textsc{HotSpot} takes the advantage over \textsc{GraalVM} and becomes more energy efficient.
In total, \textsc{HotSpot} completes $597$ requests against $510$ for \textsc{GraalVM} and $381$ for \textsc{J9}, as shown in \Cref{table:power/request}.
\textsc{HotSpot} was thus the best choice on the long term, which explains why it is always necessary to consider a warm-up phase and wait for the JIT to be triggered before evaluating the effect of the JVM or the performance of an application.
This is exactly what we did in our experiments, and why \textsc{HotSpot} was more energy efficient than \textsc{GraalVM} in \Cref{fig:all_benchs}, thus ignoring the warm-up phase would have misleading.

@@

Another measurement issue is the CPU energy variation within machines (cf. \Cref{chapter:benchmarking}), thus we executed all the comparable tests on the same node and with the recommended settings to mitigate this threat.

@@

\section{Tools and contributions}
\Cref{table:j-referral} illustrates an example of the final report of \textsf{J-Referral}.
The tool was tested for 2 real Java projects: \textsf{Zip4J}\footnote{\url{https://github.com/srikanth-lingala/zip4j}} and \textsf{K-nucleotide}.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/performance/knucleotide.html}}.
\textsf{Zip4J} runs a large file compression, while \textsf{K-nucleotide} extracts a DNA sequence, and updates a hashtable of k-nucleotide keys to count specific values.
The short report presented in \Cref{table:j-referral} shows the ratio of potential energy saving between the most and least energy consuming tested JVM (40\% and 70\% energy savings for \textsf{Zip4J} and \textsf{K-nucleotide}, respectively).
Options are available for \textsf{J-Referral} to obtain much more detailed reports including execution time, DRAM usage, split DRAM vs. CPU consumption, etc.
The tool is available as \emph{open-source software} (OSS) from our GitHub repository.\footnote{\url{https://github.com/chakib-belgaid/jreferral}}
